Estimating duration using survival analysis

### 2023-10-20 cis-perfect-testing.tex
Spellcheck

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estiamtes of the mean are sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions drive the size of the estimated tails.
In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparmeters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis cannot be independent across the hazards.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{reference discussion on extrapolation in ATACCC chapter}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior for the combination in two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-2.5cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernal density estimate smoothed) for the first 15 hazards. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

The prior on the total number of infections was centered on the true value with a dispersion parameter of 1.
This meant that the prior was very diffuse.

@@

Repeating the simulation showed negligible variation in the results (not shown).
This is caused by the large sample size (4800 detected infections).
Therefore, only results from a single simulation are shown and dicussed.

@@

The informative and vague independent priors performed similarly.
The informative prior had tighter credible intervals, as expected.
This prior was centered on the correct value for the initial part of the distribution.
Therefore, the narrow credible intervals are unsurprising.
For hazards after day 40, the informative and vague priors are the same.
Therefore, the posterior distributions are almost identical.
There remains a difference because the hazard estimates are partially dependent on earlier estimates.

>>>>

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions drive the size of the estimated tails.
In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis cannot be independent across the hazards.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see\todo{reference discussion on extrapolation in ATACCC chapter}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior for the combination in two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-2.5cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for the first 15 hazards. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
This meant that the prior was very diffuse.

@@

Repeating the simulation showed negligible variation in the results (not shown).
This is caused by the large sample size (4800 detected infections).
Therefore, only results from a single simulation are shown and discussed.

@@

The informative and vague independent priors performed similarly.
The informative prior had tighter credible intervals, as expected.
This prior was centred on the correct value for the initial part of the distribution.
Therefore, the narrow credible intervals are unsurprising.
For hazards after day 40, the informative and vague priors are the same.
Therefore, the posterior distributions are almost identical.
There remains a difference because the hazard estimates are partially dependent on earlier estimates.
